---
title: "STATISTICAL INFERENCE - Peer Assignment"
subtitle: Part 1 - Simulation Exercise
output: html_document
---
Version: V00

Date: 21-dec-2014

GitHub Repository: https://github.com/A6111E/datasciencecoursera/tree/master/Statistical_Inference/Assignment_Part1

#### Introduction:

**Exponential Distribution:** is the probability distribution that describes the time between events in a Poisson process, i.e. a process in which events occur continuously and independently at a constant average rate. 

The exponential distribution occurs naturally when describing the lengths of the inter-arrival times in a homogeneous Poisson process. Describes the time for a continuous process to change state.

In real scenarios, the assumption of a constant rate (or probability per unit time) is rarely satisfied. 

For example, the rate of incoming phone calls differs according to the time of day. But if we focus on a time interval during which the rate is roughly constant, such as from 2 to 4 p.m. during work days, the exponential distribution can be used as a good approximate model for the time until the next phone call arrives. 

Examples which yield approximately exponentially distributed variables:

- Time until a radioactive particle decays, or the time between clicks of a geiger counter. 

- Time it takes before your next telephone call

Exponential variables can also be used to model situations where certain events occur with a constant probability per unit length, such as the distance between mutations on a DNA strand, or between roadkills on a given road.

**Expected Value:** the mean or expected value of an exponentially distributed random variable $X$ with rate parameter $lambda$ is given by:

$E[X] = 1/lambda = beta$

**Variance:** $Var[X] = 1/lambda^2$

**Standard Deviation (is equal to the mean):** $sd[X] = 1/lambda = beta$

**Mean Standard Error:** $se[X] = sd/sqrt(1/n)$

**Sample Mean:** $Sample_Mean [X] = 1/sample.size(sum(xi))$

**t Confidence Interval:** $t Confidence Interval= E[X] +/- t(n-1) × sd/sqrt(n)$

*Partial Source:* http://en.wikipedia.org/wiki/Exponential_distribution

### Data Processing:
```{r Call required libraries, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
require(data.table)             # Data Table
require(ggplot2)                # Plotting
require(xtable)                 # For generating tables knitr
require(knitr)                  # Markdown
require(gridExtra)              # Grid for graphics
```

```{r knitr Setup, include = FALSE, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}

opts_chunk$set(
        fig.path = "graphs/plot_",                                         
        fig.height = 5, fig.width = 5, 
        options(scipen = 1, digits = 2),
        cache.path = "cache/",
        cache = FALSE
        )
```

```{r Clean Up Workspace, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
rm(list = ls())
gc()
```

```{r Creates Working Directories, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
dir.create("data")              # Source Data
dir.create("docs")              # Document files
dir.create("reports")           # Report files

# Other Directories: created by code (graphs / cache)
```

#### Data Set
```{r Data Set Exponential Distribution, results = "hide", echo = FALSE}
system.time({
set.seed(12345)
lambda <- 0.2                           # rate parameter Exponential Distribution
sample.size <- 40                       # sample size
nosim <- 1000                           # simulation number

# Mean - mu 
mu <- 1/lambda

# Variance - S2 
S2 <- 1/lambda^2

# Standard Deviation
sd = 1/lambda

# Standard Error
se = (1/lambda)/sqrt(sample.size)

# t Confidence Intervals - 97.5%
ciup <- mu + qt(.975, sample.size - 1) * sd / sqrt(sample.size)
cidown <- mu - qt(.975, sample.size - 1) * sd / sqrt(sample.size)

# Generates Data Tables
st <- replicate(nosim,rexp(sample.size,lambda))
p <- NULL

for (i in 1:nosim){
        p[i] <- mean(st[,i])
        }
})
```

```{r Tidy Dataset, results = "hide", echo = FALSE}
system.time({
colNames <- NULL        
for (i in 1:nosim){
        colNames[i] <- paste("Trial_", i, sep ="")
        }

# DataSet 1000 Samples x 40 Observations
DT <- as.data.table(st)
setnames(DT,colNames)

write.table(DT, "./reports/Data_Set_1000Samples_40Observations.txt", row.names = FALSE , 
            sep = ",")

# DataSet Mean for each Sample
colNames <- c("Sample_Mean")
DT1 <- as.data.table(p)
setnames(DT1,colNames)

DT1$Trial <- c(1:length(p))

write.table(DT1, "./reports/Data_Set_Mean_Sample.txt", row.names = FALSE , 
            sep = ",")
})
```

#### Table: Summary Simulation Random Exponential Distribution 
##### Simulations = 1000 - Sample Size = 40 - lambda = 0.2
```{r Summary Simulation Data Exponential Distribution 2, echo = TRUE, results = "asis", tidy = FALSE}

colNames <- c("Trial_1", "Trial_100", "Trial_500", "Trial_750", "Trial_1000")
pr1 <- DT[ , colNames, with = FALSE]

xt <- xtable(summary(pr1))
print(xt, type = "html", floating = FALSE)
```

#### Table: Mean Value for each Simulation
```{r Summary Raw Data, echo = TRUE, results = "asis", tidy = FALSE}
xt <- xtable(DT1[1:10, ], digits = 2)
print(xt, type = "html", include.rownames = FALSE, hline.after=c(1), floating = FALSE)
```

#### Graphic: Random Exponential Distribution 
```{r Random Exponential Distribution ~ Mean, echo = TRUE}
theo_center <- mu               # distribution theoretical center
cal_center <- mean(p)           # distribution calculated center

p1 <- ggplot(DT1, aes(x = Sample_Mean))
        p1 <- p1 + geom_histogram(fill = "blue",
                                binwidth = 0.25, aes(y = ..density..), colour = "black")
        p1 <- p1 + geom_vline(xintercept = theo_center, size = 0.75, colour = "red")
        p1 <- p1 + geom_vline(xintercept = cal_center, size = 0.75, colour = "green")
        mse <- round(cal_center, 3)
        p1 <- p1 + labs(title = paste("theo_center=", theo_center, "cal_center =", mse))
        p1 <- p1 + labs(subtitle = paste("red", "green"))
grid.arrange(p1, ncol = 1)
```

Remarks:

- Red Vertical Line: distribution theoretical center

- Green Vertical Line: distribution calculated center 

##### Q1 - Results:
```{r Q1-Results , echo = FALSE, tidy = FALSE}
theo_center <- mu               # distribution theoretical center
cal_center <- mean(p)           # distribution calculated center
```

The calculated center from 1000 simulations for $40$ exponential distributions with $lambda = 0.2$, is approximately $5$, which corresponds to theoretical center of distribution ($E[X] = 1/lambda = 5$). 

Increasing the simulations, the calculated gets closer to theoretical center.

**A. Theoretical Center:**              $`r theo_center`$

**B. Calculated Center:**               $`r cal_center`$

##### Q2 - Results:
```{r Q2-Results 1, echo = FALSE, tidy = FALSE}
vec_data <- as.vector(as.matrix(DT))           # all data as a vector
theo_var <- S2                                  # distribution theoretical variance
cal_var <- var(vec_data)                        # distribution calculated variance
```

The calculated variance from $1000$ simulations for $40$ exponential distributions samples  with $lambda = 0.2$ and theoretical variance $Var[X] = 1/lambda^2$ are:

**A. Theoretical Variance:**            $`r theo_var`$

**B. Calculated Variance:**             $`r cal_var`$

Increasing the simulations, the calculated gets near to theoretical variance, and the exponential distributions trend to be a standard normal distribution.

For example:

A. The calculated variance from $1$ simulation for $40$ exponential distributions random variables is $41.9$ (for this simulation).

B. The theoretical variance $Var[X] = 1/lambda^2$, is $`r theo_var`$  

Decreasing the simulations, the calculated variance gets far away to theoretical variance, and the exponential distributions trend NOT to be a standard normal distribution.

##### Q3 - Results:
```{r Q3-Results , echo = FALSE, tidy = FALSE}
cal_sd <- sd(DT1$Sample_Mean)   # distribution standard deviation
cal_center <- mean(p)           # distribution calculated center

# Sample_Means values between cal_center and +/- 1cal_sd (68%)
x <- p[p >= (cal_center - cal_sd) & p <= (cal_center + cal_sd)]
cal_sdx <- (round(length(x)/length(p),3))*100

# Sample_Means values between cal_center and +/- 2cal_sd (95%)
y <- p[p >= (cal_center - 2*cal_sd) & p <= (cal_center + 2*cal_sd)]
cal_sdy <- (round(length(y)/length(p),3))*100 

# Sample_Means values between cal_center and +/- 3cal_sd (99%)
z <- p[p >= (cal_center - 3*cal_sd) & p <= (cal_center + 3*cal_sd)]
cal_sdz <- (round(length(z)/length(p),3))*100

# t Confidence Intervals - 95%
ciup_cal <- cal_center + qt(.95, nosim - 1) * cal_sd / sqrt(nosim)
cidown_cal <- cal_center - qt(.95, nosim - 1) * cal_sd / sqrt(nosim)
```

#### Graphic: Random Exponential Distribution ~ Standard Normal Distribution 
```{r Random Exponential Distribution ~ Standard Normal, echo = TRUE}

p3 <- ggplot(DT1, aes(x = Sample_Mean))
        p3 <- p3 + geom_histogram(fill = "blue",
                                binwidth = 0.25, aes(y = ..density..), colour = "black")
        p3 <- p3 + geom_density(size = 1.25, colour = "red")
        mse <- round(cal_center, 3)
        p3 <- p3 + labs(title = paste('theo_center = ', mu, ' cal_center = ', mse))
grid.arrange(p3, ncol = 1)
```

For a standard normal distribution, $68%$, $95%$ and $99%$ of the normal density, lies within $1$, $2$ and $3$ standard deviations from the mean, respectively

The calculted percentage from the sample means of the exponential distribution ($lambda = 0.2$) are:

**A. Percentage of Values between +/-1 Standard Deviation:**         $`r cal_sdx`% / 68%$

**B. Percentage of Values between +/-2 Standard Deviation:**         $`r cal_sdy`% / 95%$

**C. Percentage of Values between +/-3 Standard Deviation:**         $`r cal_sdz`% / 99%$

Both, the graphic "Random Exponential Distribution ~ Standard Normal Distribution", and the calculated percentage values, indicates that if the sample trend to be infinite, the random exponential distribution will convert to a standard normal distribution.

The confidence interval for the quantification of uncertainty in the exponential dsitribution is:

**A. Confidence Interval 95% - Up:**                       $`r ciup_cal`$            

**B. Confidence Interval 95% - Down:**                     $`r cidown_cal`$

### Session Information
```{r Session Information, echo = FALSE}
sessionInfo()
```

```{r Delete Temporary Directories, echo = FALSE}
# Delete the '#' sign before unlink() to delete temp files
unlink("tmpDir", recursive = TRUE)
```