---
title: "INFLUENCE OF AUTOMATIC OR MANUAL TRANSMISSION ON THE CONSUMPTION IN MILES PER GALLON IN THE AUTOMOTIVE INDUSTRIE IN THE UNITED STATES"
output: html_document
---
Version: V00

Date: 25-JAN-2015

GitHub Repository: https://github.com/A6111E/datasciencecoursera/tree/master/Regression_Models

Data Source: R Data Set Package "data sets" - Data: "mtcars"

#### Synopsis:
In the attempt to predict gasoline mileage for 1973 - 1974 automobiles, road tests were preformed by Motor Trend US Magazine, in which gas mileage and 10 physical characteristics of various types of automobiles were recorded.

**Source:**
Biometrics Invited Paper: The Analysis and Selection of Variables in Linear Regression
by R. R. Hocking
Biometrics - Vol. 32, No. 1 (Mar., 1976), pp. 

**DataSet Description:** Extracted from Motor Trend US Magazine in 1.974 and comprises fuel  consumption, 10 different aspects of automobile design and performance for 32 automobiles (1973-74 models).

**- Data Frame:** $32$ observations - $11$ variables

**- Variable [mpg]:** numeric - mpg (miles per US gallon)

**- Variable [cyl]:** numeric - Number of Cylinders

**- Variable [disp]:** numeric - Cylinder Displacement (cu.in)

**- Variable [hp]:** numeric - Gross Horsepower (hp)

**- Variable [drat]:** numeric - Rear Axle ratio 

**- Variable [wt]:** numeric - Weight (lb/1000)

**- Variable [qsec]:** numeric - 1/4 Mile Time (sec)

**- Variable [vs]:** numeric - Cylinder Configuration (V - S: straight ) (0 = V, 1 = S)

**- Variable [am]:** numeric - Transmission (0 = automatic, 1 = manual)

**- Variable [gear]:** numeric - Number of forward Gears

**- Variable [carb]:** numeric - Number of Carburetors

With this data set, this research tries to answer:

1. Is an automatic or manual transmission better for MPG (gas mileage)?

2. Quantify the MPG (gas mileage) difference between automatic and manual transmissions

### Executive Summary:

- After a Exploratory Analysis and Model Finding, the best linear model, that predict the gasoline mileage for this data set is:

$$mpg = 9.62 - 3.92wt + 1.23qsec + 2.94am$$

- An Adjusted $R^2 = 0.83$, indicates that **$83.36%$** of the variance in the outcome variable $mpg$ can be explained by the predictors variables $wt, qsec, am$ and this linear model.

- The **Manual Transmission** is more efficient than the **Automatic Tranmission** in $2.94 miles/gal$

### Exploratory Analysis:
```{r Call required libraries, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
library(datasets)               # Different DataSets
library(data.table)             # Extension of data frame
library(ggplot2)                # Plotting
library(xtable)                 # For generating tables knitr
library(knitr)                  # Markdown
library(gridExtra)              # Grid for graphics
library(plyr)                   # Tools for splitting, applying and combining data
library(GGally)                 # Extension to ggplot2
library(car)                    # Companion to Applied Regression
library(gtools)                 # Various R Programming Tools
```

```{r knitr Setup, include = FALSE, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}

opts_chunk$set(
        fig.path = "graphs/plot_",                                         
        fig.height = 5, fig.width = 10, 
        options(scipen = 1, digits = 2),
        cache.path = "cache/",
        cache = FALSE
        )
```

```{r Clean Up Workspace, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
rm(list = ls())
gc()
```

```{r Creates Working Directories, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
dir.create("data")              # Source Data
dir.create("docs")              # Document files
dir.create("reports")           # Report files
dir.create("code")           # Code files

# Other Directories: created by code (graphs / cache)
```

```{r Load DataSet, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
system.time({
mtcars <- as.data.table(mtcars)        # Source: DataSet R Package
list2env(mtcars , envir = .GlobalEnv)

write.table(mtcars, "./reports/mtcars.txt", row.names = FALSE , 
            sep = ",")

# Summary Mileage per US Gallon - per Trasnmission Type (Manual = 1 / Automatic = 0)
mpg_transType <- subset(mtcars, select = c("mpg", "am"))
write.table(mpg_transType, "./reports/Mileage_per_US_Gallon_TType.txt", row.names = FALSE , sep = ",")
})
```

```{r Data Class, dependson="loadData", echo = FALSE, results = "hide"}
sapply(mtcars[1,],class)
sum(is.na(mtcars))
```

#### Table 1: Data Preview 
```{r Data Preview, echo = FALSE, results = "asis", tidy = FALSE}
xt <- xtable(head(mtcars))
print(xt, type = "html", floating = FALSE)
```

```{r Exploratory Data_1, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
g1 <- ggpairs(mtcars, columns = c(1:11), diag = list(continuous = "density", params = c(colour="red")), upper = list(continuous = "cor", params = c(size = 3.5), axisLabels= "show"), title = "", lower=list(continuous="smooth", params=c(colour="blue")))

# Putting Axis Text Blank
g1 <- g1 + theme(axis.text.y = element_blank(), axis.title.y = element_blank())
g1 <- g1 + theme(axis.text.x = element_blank(), axis.title.x = element_blank())

# Box Plot Mileage per US Gallon ~ Automatic / Manual Transmission
g2 <- ggplot(data = mtcars, aes(y = mpg, x = am, fill = factor(am)))
g2 <- g2 + geom_boxplot(colour = "black", size = 0.8)
g2 <- g2 + xlab("Transmission Type: 0: Automatic 1: Manual") + ylab("Miles per US Gallon")

```

```{r Exploratory Data_2, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
loopLength <- length(sapply(mtcars[1,],class)) - 1
sta_coeff <- NULL
        
for (i in 1:loopLength) {
        j <- 1 + i
        
        # Linear Model
        fit <- lm(paste("mtcars$mpg ~ mtcars$", colnames(mtcars)[j], sep=""))
        sumCoef <- summary(fit)$coefficients    # Coefficients Summary
        sumCorr <- summary(fit, correlation = TRUE)$correlation         # Correlation
        
        beta0 <- sumCoef[1]     # Intercept
        beta1 <- sumCoef[2]     # Slope
        
        R2 <- summary(fit)$r.squared    # R2

        adjR2 <- summary(fit)$adj.r.squared     # Adjusted R2
        
        # Correlation
        y <- as.vector(subset(mtcars, select = "mpg"))
        x <- as.vector(subset(mtcars, select = colnames(mtcars)[j]))
        
        corr <- cor(y, x)    #Correlation
        
        sig <- summary(fit)$sigma    # Sigma
        
        sta_coeff <- rbind(sta_coeff, c(colnames(mtcars)[j], round(beta0,3), 
                                round(beta1,3), round(R2, 3), round(adjR2, 3), 
                                round(corr, 3), round(sig, 3)))
                }        
```
 
```{r Exploratory Data_3, echo = FALSE, results = "asis", tidy = FALSE}
sta_coeff <- as.data.frame(sta_coeff)

# Changing Columns from Factor to Numeric
sta_coeff$V2 = as.numeric(as.character(sta_coeff$V2))
sta_coeff$V3 = as.numeric(as.character(sta_coeff$V3))
sta_coeff$V4 = as.numeric(as.character(sta_coeff$V4))
sta_coeff$V5 = as.numeric(as.character(sta_coeff$V5))
sta_coeff$V6 = as.numeric(as.character(sta_coeff$V6))
sta_coeff$V7 = as.numeric(as.character(sta_coeff$V7))

# Changing Column Names
colNames <- c("Variable", "Intercept_Beta0", "Slope_Beta1", "R^2", "Adjusted_R2",
              "Correlation", "Sigma")

setnames(sta_coeff,colNames)

# Sort by Correlation
# sapply(sta_coeff[1,],class)     # Class Test

sta_coeff <- sta_coeff[with(sta_coeff, order(-abs(Correlation))), ]
```

```{r Exploratory Data_4, echo = FALSE, results = "asis", tidy = FALSE}
# Testing Results

fit1 <- lm(mtcars$mpg ~ mtcars$cyl)

sumCoef <- summary(fit1)$coefficients    # Coefficients Summary
sumCorr <- summary(fit1, correlation = TRUE)$correlation         # Correlation
        
beta0 <- sumCoef[1]     # Intercept
beta1 <- sumCoef[2]     # Slope
        
R2 <- summary(fit1)$r.squared    # R2
        
corr <- sumCorr[2,1]    #Correlation

corr <- cor(mtcars$mpg, mtcars$cyl)

sig <- summary(fit1)$sigma    # Sigma
```

```{r Exploratory Data_5, echo = FALSE, results = "asis", tidy = FALSE}
# Creates Variable Names according with Correlation descending order
vector <- as.list(sta_coeff[,"Variable"])

names(vector) <- paste("var", 1:length(vector), sep = "")

list2env(vector , envir = .GlobalEnv)
```

```{r Exploratory Data_6, echo = FALSE, results = "asis", tidy = FALSE}
# Creates LineAr Model Combinations
vars <- as.vector(colnames(mtcars))

N <- as.list(1:(length(vars)-1))
COMB <- sapply(N, function(m) combn(x = vars[2:length(vars)], m))
COMB2 <- list()
k <- 0

for(i in seq(COMB)){
    tmp <- COMB[[i]]
    for(j in seq(ncol(tmp))){
        k <- k + 1
        COMB2[[k]] <- formula(paste("mpg", "~", paste(tmp[,j], collapse = " + ")))
    }
}

possComb <- as.data.frame(matrix(unlist(COMB2), nrow = length(COMB2), byrow = TRUE))
colNames <- c("Combinations")
setnames(possComb,colNames)

combNum <- length(COMB2)
```

- The value of cylinder is a categorical variable

- Half of the cars get 19.20 miles per gallon or less (Median). 

- The variability or spread for $mpg$ is equal to $\sigma = 6.02$

- For the multivariate $mpg$ data, there should be some relationship with the size of the engine (# of cylinders, gross horsepower or even the cylinder displacement). 

- Plotting $mpg$ ~ $cyl$, a decreasing trend on $mpg$ as the # of $cyl$ increases is obvious, and it should be a candidate fit a regression line.

- The summary on table "Table A1: Statistical Data Summary - Mileage per US Gallon " included on the Appendix, shows statistical data like the mean, standard deviation, variance, median, maximal and minimal values, for each type of transmission (Automatic - Manual) in relationship with the Mileage per US Gallon (mpg).

- According with this table, the maximal and minimal values for **Manual Transmission**, are higher than corresponding values for **Automatic Transmission**, although its standard deviation is greater.

- Taking "Mileage per US Gallon (mpg)" as outcome, and using the other variables as predictors, statistical coefficients will be calculated for a linear model during a  exploratory analysis.

- This can be observed on Graphic A1 included on the Appendix.

**Initial Conclusions:**

- **Manual Transmission**: higher throughput for the outcome "Mileage per US Gallon (mpg)" in compassion with the **Automatic Transmission**.

- **Transmision Type (am)**: according with the Statistic Coefficients Table and related graphic (please see the Appendix), the **am** variable is the **7th** place of influence on the **mpg**  variable.

- According with the correlation (please see Appendix - Table A2) for the required outcome $mpg$ and the predictors, the impact on it in descending order is: $1. `r var1`, 2. `r var2`, 3. `r var3`, 4. `r var4`, 5. `r var5`, 6. `r var6`, 7. `r var7`, 8. `r var8`, 9. `r var9`, 10. `r var10`$

- The variable $qsec$ is the required time for driving $/1/4$ of mile. From physics  in general, less time means more acceleration and final speed and more gas consumption. Although this variable has the smallest correlation with $mpg$, it can not be avoided from the model analysis.  

- The model with $mpg$ as outcome, should **NOT** have **ONLY** one predictor.

- The number of possible combinations for this linear model is $`r combNum`$.

- Due to the high number of possible combinations, it's necessary to use an algorithm (Stepwise Algorithm, where the best model corresponds to the smallest AIC).

### Statistical Inference

**Hypothesis 1**:

- Null hypothesis $H_{01}$: is that the residuals of the calculated linear model are normally distribute and do not have trends.

- Alternative Hypothesis $H_{a1}$: opposite as the null hypothesis.

**Hypothesis 2**:

- Null hypothesis $H_{02}$: is that the base model $mpg$ ~ $am$ (trying to answer the  main questions), is less representative that the best model ($mpg$ as outcome and several variables as predictors).

- Alternative Hypothesis $H_{a2}$: opposite as the null hypothesis.

### Modeling

- Including all variables in an initial linear model, all p-value´s are not less than 0.05, and a linear model including all variables **is not** representative.     

#### Table 2: Summary Linear Model - All Variables
```{r Summary Linear Model, echo = FALSE, results = "asis", tidy = FALSE}
xt <- xtable(summary(lm(mpg ~ . , data = mtcars))$coefficients) 
print(xt, type = "html", floating = FALSE)
```

### Linear Model Selection
```{r Model_1, echo = FALSE, results = "hide", tidy = FALSE}
initialmodel <- lm(mpg ~ ., data = mtcars)
bestmodel <- step(initialmodel, direction = "both")
```

```{r Model_2, echo = FALSE, tidy = FALSE}
R2 <- summary(bestmodel)$r.squared      # R2
adjR2 <- summary(bestmodel)$adj.r.squared       # Adjusted R2
adjR2per <- adjR2 * 100         # Adjusted R2 in %
remadjR2per <- 100 - adjR2per   # Remainder Adjusted R2
sigma <- summary(bestmodel)$sigma       # Sigma

sumCoef <- summary(bestmodel)$coef

# Betas
beta0 <- sumCoef[1]     # Intercept
beta1 <- sumCoef[2]     # slope for Var2 (wt)
beta2 <- sumCoef[3]     # slope for Var3 (qsec)
beta3 <- sumCoef[4]     # slope for Var4 (am)

# Variable Names
var1 <- names(bestmodel$model)[1]
var2 <- names(bestmodel$model)[2]
var3 <- names(bestmodel$model)[3]
var4 <- names(bestmodel$model)[4]

# p_values
p_value_var2 <- format(sumCoef[2,4], scientific = TRUE, digits = 3)
p_value_var3 <- format(sumCoef[3,4], scientific = TRUE, digits = 3)
p_value_var4 <- format(sumCoef[4,4], scientific = TRUE, digits = 3)
```

```{r Model_3, echo = FALSE, results = "hide", tidy = FALSE}
# Comparassion between Real Values and Calculated Values
est_data <- as.data.table(mtcars)
est_data <- est_data[ ,c("cyl", "disp", "hp", "drat", "vs", "gear", "carb") := NULL]

est_data$mpg_estimated <- round(beta0 + beta1*est_data$wt + beta2*est_data$qsec 
                          + beta3*est_data$am, 3)

est_data$residuals <- round(resid(bestmodel), 3)

# Maximal Residual
e <- resid(bestmodel)         # Calculates the residual
maxResValue <- max(e) 
minResValue <- min(e) 
```

The best fitting model for the data set has the following statistical data:

- $R^2$ = `r R2`

- Adjusted $R^2$ = `r adjR2`

- $\sigma$ = `r sigma`

- Linear Model: $`r var1` = `r beta0` `r beta1``r var2` + `r beta2``r var3` + `r beta3``r var4`$

- An Adjusted $R^2$ = `r adjR2` (maximal value obtained for the best fitting linear model) value, indicates that `r adjR2per`% of the variance in the outcome variable $mpg$ can be explained by the predictors variables: $`r var2`$, $`r var3`$, $`r var4`$. 

- The remaining `r remadjR2per`% can be attributed to unknown, hidden or not included  variables or inherent variability.

- The p_values values are all representative:

a. $pvalue(`r var2`) = `r p_value_var2`$

b. $pvalue(`r var3`) = `r p_value_var3`$

c. $pvalue(`r var4`) = `r p_value_var4`$

- The maximal / minimal residual values are:

a. Maximal: $`r maxResValue`$

b. Minimal: $`r minResValue`$ 

#### Statistical Inference Conclusions

- The Null hypothesis $H_{01}$ *could not be rejected* (residuals of the calculated linear model are normally distribute and do not have trends).

```{r Model_4, echo = FALSE, results = "hide", tidy = FALSE}
# Anova
basemodel <- lm(mpg ~ am, data = mtcars)
pvalue <- anova(bestmodel, basemodel)[2,6]
```

- The Null hypothesis $H_{02}$ *could be rejected* (base model $mpg$ ~ $am$ is **NOT** representative to estimate the $mpg$).

#### Final Analysis:

- For defining which transmission, automatic or manual, is better for MPG (gas mileage) we have:

(1) **Variable [am]:** numeric - Transmission (0 = automatic, 1 = manual)

(2) **Slope [am]:** $`r beta3`$

(3) **Automatic Transmission =** $`r beta3`* 0 = 0$
replacing:
$$`r var1` = `r beta0` `r beta1``r var2` + `r beta2``r var3`$$

(4) **Manual Transmission =** $`r beta3`* 1 = `r beta3`$
replacing:
$$`r var1` = `r beta0` `r beta1``r var2` + `r beta2``r var3` + `r beta3``r var4`$$

(5) **Findings:** taking $wt$ and $qsec$ as constant values, the **Manual Transmission** adds $`r beta3`$ $miles/gal$ in comparison with the **Automatic Tranmission**

(6) In other words, the **Manual Transmission** is more efficient than the **Automatic**.

- For quantifying the MPG (gas mileage) difference between automatic and manual transmissions we have:

(1) **Variable [am]:** numeric - Transmission (0 = automatic, 1 = manual)

(2) **Slope [am]:** $`r beta3`$

(3) **Automatic Transmission =** $`r beta3` * 0 = 0$

(4) **Manual Transmission =** $`r beta3`* 1 = `r beta3`$

(5) **Findings:** the MPG (gas mileage) difference between automatic and manual transmissions is $`r beta3`$ $miles/gal$

### Appendix
```{r Appendix_0, echo = FALSE, results = "hide", tidy = FALSE}
# mpg
mean(mpg)
summary(mpg)
sd(mpg)
```

#### Graphic 0: Mean $mpg$ ~ $Cylinder$ 
```{r Appendix_01, echo = FALSE, results = "asis", tidy = FALSE}
# Cylinder
mm <- ddply(mtcars, "cyl", summarise, mmpg = mean(mpg))
mm$cyl <- as.factor(mm$cyl)

g0 <- ggplot(mtcars, aes(x = factor(cyl), y = mpg)) + geom_point(size = 5, 
        colour = "red") + xlab("Cylinder") + ylab("Mean MPG")
g0 <- g0 + geom_line(data = mm, aes(x = cyl, y = mmpg, group =1))
g0
```

#### Table A1: Statistical Data Summary - Mileage per US Gallon 
```{r Appendix_1, echo = FALSE, results = "asis", tidy = FALSE}
# Summary Mileage per US Gallon
xt <- xtable(ddply(mtcars,.(am), summarize, Average = mean(mpg), Std_Deviation = sd(mpg), Variance = var(mpg), Median = median(mpg), Max = max(mpg),  Min = min(mpg)))
print(xt, type = "html", floating = FALSE)
```

Remarks:

- Transmission Type (am):  1 (Manual Transmission) - 0 (Automatic Transmission)

- Average / Std_Deviation / Variance: statistic data for Mileage per US Gallon (mpg) 

- GitHub Repository: /reports

#### Graphic 1: Motor Trend Car Road Tests - Linear Model 
```{r Appendix_2, echo = FALSE, results = "hide", warning = FALSE, message = FALSE}
g1
grid.arrange(g2, nrow = 1, ncol = 2)
```

#### Table A2: Statistic Coefficients
```{r Appendix_3, echo = FALSE, results = "asis", warning = FALSE, message = FALSE}
print(xtable(sta_coeff), type = "html", floating = FALSE)

write.table(sta_coeff, "./reports/Statistical_Data_Coefficients.txt", row.names = FALSE , sep = ",")
```

#### Table A3: Summary Best Linear Model 
```{r Appendix_4, echo = FALSE, results = "asis", tidy = FALSE}
print(xtable(summary(bestmodel)), type = "html", floating = FALSE)

write.table(xtable(summary(bestmodel)), "./reports/Summary_BestModel.txt", row.names = FALSE , sep = ",")
```

#### Table A4: Summary Anova - Best Linear Model 
```{r Appendix_5, echo = FALSE, results = "hide", tidy = FALSE}
# Anova
basemodel <- lm(mpg ~ am, data = mtcars)
anova(basemodel, bestmodel)
```

```{r Appendix_5_1, echo = FALSE, results = "asis", tidy = FALSE}
print(xtable(anova(basemodel, bestmodel)), type = "html", floating = FALSE)

write.table(xtable(anova(basemodel, bestmodel)), "./reports/Anova_BestModel.txt", row.names = FALSE , sep = "")
```

#### Graphic A2: Best Linear Model - Residuals
```{r Appendix_6, echo = FALSE, results = "asis", warning = FALSE, message = FALSE}
ggplotRegression <- function (fit) {
  ggplot(fit$model, aes_string(x = names(fit$model)[2], y = names(fit$model)[1])) + 
  geom_point(size = 2, colour = "red", alpha = 1) +  
  stat_smooth(method = "lm", col = "red") + 
  geom_point(data = mtcars, aes(wt, mpg), size = 4, colour = "black", alpha = 0.4)
  }

# Plotting
g3 <- ggplotRegression(bestmodel)
                  
lmodel <- paste(var1, "=", round(beta0,3), round(beta1,3), var2, "+", 
                round(beta2,3), var3, "+", round(beta3,3), var4, sep = "")

p_values <- paste(var2, "=", format(p_value_var2, scientific = TRUE, digits = 3), 
                  ";",  
                  var3, "=", format(p_value_var3, scientific = TRUE, digits = 3), 
                  ";", 
                  var4, "=", format(p_value_var4, scientific = TRUE, digits = 3), 
                  sep = "")

plotTitle <- as.character(lmodel)
plotSubTitle1 <- as.character(c(paste("p_values:", p_values)))
plotSubTitle2 <- as.character(c(paste("Adj R2=", round(adjR2, 3))))

g3 <- g3 + labs(title = c(paste(plotTitle, " \n ",  plotSubTitle1, " \n ", 
                        plotSubTitle2, " \n "))) 

g3 <- g3 + annotate("text", label = "Original Data", x = 2.5, y = 4, size = 5, colour = "black")

g3 <- g3 + annotate("text", label = "Estimated Data", x = 2.5, y = 2, size = 5, colour = "red")

# Residuals mpg
e <- resid(bestmodel)         # Calculates the residual

g4 <- ggplot(mtcars, aes(x = mpg, y = e))
g4 <- g4 + xlab("mpg")
g4 <- g4 + ylab("Residual mpg")
g4 <- g4 + geom_hline(yintercept = 0, size = 2)
g4 <- g4 + geom_point(size = 4, colour = "black", alpha = 0.4)
g4 <- g4 + geom_point(size = 2, colour = "red", alpha = 1)
g4 <- g4 + ggtitle("mpg - Residuals")

# Residuals Histogramm - Normally Distributed
g5 <- ggplot(data = NULL, aes(x = e))
g5 <- g5 + geom_histogram(fill = "salmon", binwidth = 1, aes(y = ..density..), 
                        colour = "black")
g5 <- g5 + geom_density(size = 1) + ggtitle("Histogram Residuals")

# Residuals is on a straight line
g6 <- ggplot(data = as.data.frame(qqnorm( e, plot=F)), mapping=aes(x=x, y=y)) + 
    geom_point(size = 5) + geom_smooth(method="lm", se=FALSE, size = 1) + ggtitle("mpg - Residuals") + 
    xlab("Theoretical Quantiles") + ylab("Sample Quantiles")

grid.arrange(g3, g4, nrow = 1, ncol = 2)

grid.arrange(g5, g6, nrow = 1, ncol = 2)
```

#### Table A5: Estimated Data vs Real Data 
```{r Appendix_7, echo = FALSE, results = "asis", tidy = FALSE}
xt <- xtable(head(est_data))
print(xt, type = "html", floating = FALSE)

write.table(est_data, "./reports/Estimated_Data_vs_Calculated_Datas.txt", row.names = FALSE , sep = ",")
```

### Session Information
```{r Session Information, echo = FALSE}
sessionInfo()
```

```{r Delete Temporary Directories, echo = FALSE}
# Delete the '#' sign before unlink() to delete temp files
unlink("tmpDir", recursive = TRUE)
```